Experiences with Planning for Natural Language Generation

### 2009-06-03 trunk/Papers/ci-spark/experiments.tex
done with my experiments

We converted these generation problem instances into planning problem
instances as described in Section~\ref{sec:domains}, and then ran
several different planners on them. We used three off-the-shelf
planners: FF 2.3 \citep{HoffmannNebel01}, Metric-FF
\citep{hoffmann:ecai-02}, and SGPLAN 6
\citep{hsu06:_new_featur_in_sgplan_for}; all of these were highly
successful at the recent IPC competitions, and unlike many other IPC
participants, support a fragment of PDDL with quantified and
conditional effects, which is necessary in our domain. In addition, we
used an ad-hoc implementation of GraphPlan \citep{Blum1997} written in
Java; unlike the three off-the-shelf planners, it only computes
instances of literals and operators as they are needed in the course
of the plan search, instead of computing all ground instances in a
separate preprocessing step. Finally, we reimplemented the incomplete
greedy search algorithm used in the SPUD system \citep{Stone2003a}.

The results of this experiment are shown in the graph in
Figure~\ref{fig:runtimes-crisp}.\footnote{All runtimes in
  Sections~\ref{sec:exper-1:-sent} and \ref{sec:exper-2:-sent-xtag}
  were measured on a single core of an AMD Opteron 8220 CPU running at
  2.8 GHz, under Linux. FF 2.3 and Metric-FF were recompiled as 64-bit
  binaries. Java programs were executed under Java 1.6.0\_13 in 64-bit
  mode. They were allowed to ``warm up'', i.e. the JVM was given the
  opportunity to just-in-time compile the relevant bytecode by running
  the planner three times and discarding the runtimes before taking
  the actual measurements. All runtimes are averaged over three runs
  of the planners.} The input parameters $(m,n)$ are plotted in
lexicographic order on the horizontal axis, and the runtime is shown
in seconds on the vertical axis, on a logarithmic scale. These results
reveal a number of interesting insights. First, the search times of FF
and Metric-FF (shown as thinner lines) significantly outperform
SGPLAN's search in this domain---on the largest instances, by a factor
of over 100.\footnote{For FF and Metric-FF, we report the
  ``searching'' and ``total'' times reported by the planners. For
  SGPLAN, we report the ``total'' time and the difference between the
  ``total'' and ``parsing'' times.} Second, FF and Metric-FF perform
very similarly to each other, and their search times are almost the
same as those of the SPUD algorithm, which is impressive because they
are complete search algorithms, whereas SPUD's greedy algorithm is
not.

@@

The first experiment already gives us some initial insights into the
appropriateness of planning for the sentence generation domain: On the
examples we looked at, search times were low, but FF and SGPLAN spend
a lot of time on the initial grounding step.  However, one weakness of
this experiment is that it uses a tiny grammar, consisting of just
those 12 lexicon entries that are needed for the experiment. While the
grounding problem can only get worse with larger grammars, the
experiment by itself does not allow us to make clear statements about
the search efficiency.

To address this problem, we ran a second sentence generation
experiment. This time, we used the XTAG Grammar \citep{xtag01:_tr}, a
large-scale TAG grammar of English. XTAG contains lexicon entries for
about 17,000 uninflected words using about 1100 different elementary
trees. Although XTAG does not contain semantic information, it is
possible to automatically equip the lexicon entries with inferred
semantic representations based on the words in the lexicalized
elementary tree. The result is a highly ambiguous grammar: The most
ambiguous word, ``ask'', is the anchor of 314 lexicon entries.

@@

Fig.~\ref{fig:xtag-graph} reports the runtimes we measured in this
experiment for FF. Because the computations in this experiment were
very memory-intensive, we recompiled FF 2.3 as a 64-bit binary. We
also ran the experiments with a 64-bit version of Metric-FF; although
our planning problem is purely sequential and not metric, Metric-FF
avoids a bug which caused FF 2.3 to switch to a best-first search
strategy beyond a certain point; the jump in runtime is clearly
visible in the figure. We did not run this experiment with SGPLAN
because we didn not have access to a 64-bit version of either SGPLAN
5.2.2 or SGPLAN 6; some initial experiences with the 32-bit version
indicate that SGPLAN switches to FF's best-first strategy even for
small problem instances, and thus would be slower than the versions of
FF we reported here. We also do not report runtimes for our Java
implementation of Graphplan, because it was unusuably slow for serious
problem instances: For $k=1$ and $n=3$, it already took over two
minutes, and it exceeded its memory limit of 16 GB for
$n>3$. \todo{compare to SPUD reconstruction here}

There are a number of observations we can make in this
experiment. First of all, the experiment confirms that FF's Enforced
Hill-Climbing search strategy works very well for the sentence
generation task: Although we are now generating with a large grammar,
FF generates a 39-word sentence ($k=1,n=6$) in under a second of
search time. This efficiency is a direct result of this particular
search strategy: Whenever FF falls back to the best-first strategy,
runtimes become much slower (compare the graphs for FF 2.3 and
Metric-FF). Second, FF's runtime is still dominated by the
preprocessing stage. For instance, Metric-FF spends about 10 seconds
on search for $k=1,n=10$, compared to its total runtime of about 65
seconds. This effect becomes more pronounced as we increase $k$: For
$k=2$, we reach 65 seconds of total runtime at $n=4$, but here
Metric-FF only spends about a second on search. For $k=3$, neither FF
nor Metric-FF were able to solve any of the input
instances. \todo{bla} This is to be expected, given that the planning
operators for the verbs have $k+2$ parameters (see
Fig.~\ref{fig:white-rabbit-as-planning}), and thus the number of
action instances grows by a factor of the universe size every time we
increase $k$ by one. A planner which computes all ground instances of
the operators thus takes exponential time in $k$ for preprocessing.

>>>>

We converted these generation problem instances into planning problem
instances as described in Section~\ref{sec:domains}, and then ran
several different planners on them. We used three off-the-shelf
planners: FF 2.3 \citep{HoffmannNebel01}, Metric-FF
\citep{hoffmann:ecai-02}, and SGPLAN 6
\citep{hsu06:_new_featur_in_sgplan_for}; all of these were highly
successful at the recent IPC competitions, and unlike many other IPC
participants, support a fragment of PDDL with quantified and
conditional effects, which is necessary in our domain. In addition, we
used an ad-hoc implementation of GraphPlan \citep{Blum1997} written in
Java; unlike the three off-the-shelf planners, it only computes
instances of literals and operators as they are needed in the course
of the plan search, instead of computing all ground instances in a
separate preprocessing step. Finally, we reimplemented the incomplete
greedy search algorithm used in the SPUD system \citep{Stone2003a} in
Java.

The results of this experiment are shown in the graph in
Figure~\ref{fig:runtimes-crisp}.\footnote{All runtimes in
  Sections~\ref{sec:exper-1:-sent} and \ref{sec:exper-2:-sent-xtag}
  were measured on a single core of an AMD Opteron 8220 CPU running at
  2.8 GHz, under Linux. FF 2.3 and Metric-FF were recompiled as 64-bit
  binaries and run with a memory limit of 32 GB. Java programs were
  executed under Java 1.6.0\_13 in 64-bit mode. They were allowed to
  ``warm up'', i.e. the JVM was given the opportunity to just-in-time
  compile the relevant bytecode by running the planner three times and
  discarding the runtimes before taking the actual measurements. All
  runtimes are averaged over three runs of the planners.} The input
parameters $(m,n)$ are plotted in lexicographic order on the
horizontal axis, and the runtime is shown in seconds on the vertical
axis, on a logarithmic scale. These results reveal a number of
interesting insights. First, the search times of FF and Metric-FF
(shown as thinner lines) significantly outperform SGPLAN's search in
this domain---on the largest instances, by a factor of over
100.\footnote{For FF and Metric-FF, we report the ``searching'' and
  ``total'' times reported by the planners. For SGPLAN, we report the
  ``total'' time and the difference between the ``total'' and
  ``parsing'' times.} Second, FF and Metric-FF perform very similarly
to each other, and their search times are almost the same as those of
the SPUD algorithm, which is impressive because they are complete
search algorithms, whereas SPUD's greedy algorithm is not.

@@

The first experiment already gives us some initial insights into the
appropriateness of planning for the sentence generation domain: On the
examples we looked at, the search times were quite acceptable, but FF
and SGPLAN spent a lot of time on the initial grounding step.
However, one weakness of this experiment is that it uses a tiny
grammar, consisting of just those 12 lexicon entries that are needed
for the experiment. While the grounding problem can only get worse
with larger grammars, the experiment by itself does not allow us to
make clear statements about the search efficiency.
To address this problem, we ran a second sentence generation
experiment. This time, we used the XTAG Grammar \citep{xtag01:_tr}, a
large-scale TAG grammar of English. XTAG contains lexicon entries for
about 17,000 uninflected words using about 1100 different elementary
trees. Although XTAG does not contain semantic information, it is
possible to automatically equip the lexicon entries with inferred
semantic representations based on the words in the lexicalized
elementary trees. The result is a highly ambiguous grammar: The most
ambiguous word, ``ask'', is the anchor of 314 lexicon entries.

@@

Fig.~\ref{fig:xtag-graph} reports the runtimes we measured in this
experiment for FF, Metric FF, and the SPUD reimplementation. We do not
report runtimes for SGPLAN, because we could not recompile SGPLAN as a
64-bit binary, and the 32-bit version ran out of memory very
quickly. We also do not report runtimes for our Java implementation of
Graphplan, because it was unusuably slow for serious problem
instances: For $k=1$ and $n=3$, it already took over two minutes, and
it exceeded its memory limit of 16 GB for $n>3$. This may be a
limitation of our naive implementation rather than the GraphPlan
algorithm itself.

Nonetheless, there are a number of observations we can make in this
experiment. First of all, the experiment confirms that FF's Enforced
Hill-Climbing search strategy works very well for the sentence
generation task: Although we are now generating with a large grammar,
FF generates a 39-word sentence ($k=1,n=6$) in under a second of
search time. This level of efficiency is a direct result of using this
particular search strategy: For $k=1$ and $n>6$, FF 2.3 (but not
Metric-FF) fell back to the best-first search strategy, which causes a
dramatic loss of search effiency. It is also encouraging that
Metric-FF still performs comparably to SPUD in terms of pure search
time. We believe that by FF's technique of evaluating actions by
estimating the distance to a goal state for the relaxed problem
essentially picks out the same evaluation function as SPUD's domain-specific
heuristic, and the enforced hill-climbing strategy needs to backtrack
very little in this domain and thus performs similarly to SPUD's
greedy search. However, SPUD's incompleteness manifests itself in this
experiment by its inability to find any plan for $k>1$ and $n>1$,
whereas FF and its variants still (correctly) find these plans.

Second, FF's runtime is still dominated by the preprocessing
stage. For instance, Metric-FF spends about 10 seconds on search for
$k=1,n=10$, compared to its total runtime of about 65 seconds. This
effect becomes more pronounced as we increase $k$: For $k=2$, we reach
65 seconds of total runtime at $n=4$, but here Metric-FF only spends
about a second on search. For $k=3$, neither FF nor Metric-FF were
able to solve any of the input instances within their memory
limit. This is consistent with the observation that the planning
operators for the verbs have $k+2$ parameters (see
Fig.~\ref{fig:white-rabbit-as-planning}), and thus the number of
action instances grows by a factor of the universe size every time we
increase $k$ by one. A planner which computes all ground instances of
the operators thus takes exponential time in $k$ for preprocessing.
